---
title: "Quant: Assignment 1 + 2"
author: "Samantha Page + Lamei Zhang"
date: "11/2/2021"
output: 
  html_document:
    theme: cerulean
    toc: true
    toc_depth: 6
    toc_float: true
---
```{r setup, include=FALSE, echo = FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo = FALSE, message=FALSE}
library(tidyverse)
library(tidycensus)
library(readxl)
library(knitr)
library(sf)
library(gridExtra)
```

# Research Question

How does affordable housing impact property value? Specifically for the purposes of this assignment, how does affordable housing (defined a variety of ways) affect property values in San Francisco, California, between 2013-14 and 2018-19?

## Prior Research 

In a policy brief written by the Center for Housing Policy (n.d.), findings suggest that affordable housing generally does not have significant impacts on surrounding property values, though the effects vary based on specific context. For example, clustered developments showed some negative impact on housing developments, while more diffuse development of physical affordable housing structures had little to no impact. Nguyen (2005) confirms these findings in her literature review "Does Affordable Housing Detrimentally Affect Property Values?" Nguyen generally concludes that when affordable housing is developed in depressed neighborhoods and when the quality of the housing is poor, the values of proximate property are shown to decline. Conversely, the review found no effect of affordable housing on property values when developed to good quality standards and in healthy neighborhoods. Nguyen advocates for more studies to reach more conclusive understandings.

# Data

The sample population for this study is property values in San Francisco, California, in 2013-14 and 2018-19. The analysis will include the following variables: 

* San Francisco property values (2013-2014)
* San Francisco property values (2018-2019)
* Addresses/Sites of affordable housing developments
* Addresses/Sites where public benefits are being used for affordable housing (i.e. LIHTC)
* Total number of subsidized units at each site
* Income of household and income as percent of AMI
* Percent change of property value from 2013-2014 to 2018-2019

## Load Data 

``` {r message=FALSE}
sf_1314 <- read_csv("sf_tax_1314.csv") 

names(sf_1314)[1] <- 'Site'

names(sf_1314)[3] <- 'Block_Lot'

names(sf_1314)[31] <- 'LandValue_1314'

names(sf_1314)[32] <- 'PropValue_1314'

sf_1314_clean <- sf_1314 %>%
  select(Site, Block_Lot, LandValue_1314, PropValue_1314) %>%
    mutate(TotValue_1314 = PropValue_1314 + LandValue_1314)

sf_1819 <- read_csv("sf_tax_1819.csv") 

names(sf_1819)[1] <- 'Site'

names(sf_1819)[2] <- 'Block_Lot'

names(sf_1819)[31] <- 'LandValue_1819'

names(sf_1819)[32] <- 'PropValue_1819'

sf_1819_clean <- sf_1819 %>%
  select(Site, Block_Lot, LandValue_1819, PropValue_1819) %>%
    mutate(TotValue_1819 = PropValue_1819 + LandValue_1819) 

affhousing_loc <- read_csv("policymap_AH.csv")

names(affhousing_loc)[10] <- 'ZIPCode'

names(affhousing_loc)[12] <- 'TotalSubUnits'

names(affhousing_loc)[13] <- 'No_residents'

names(affhousing_loc)[38] <- 'AvgIncome'

names(affhousing_loc)[41] <- 'Income_pct_AMI'
```

## Determining Percent Change of Property Value and Joining Data

```{r, message=FALSE}
data <- left_join(sf_1314_clean, sf_1819_clean) %>%
  left_join(affhousing_loc) %>%
  mutate(pct_change = ((TotValue_1314 - TotValue_1819)/TotValue_1314)*-100) %>%
  mutate(newpropval = ifelse(pct_change > 0.0, "Increase", "Decrease")) %>%
  select(Site, Block_Lot, ZIPCode, TotalSubUnits, AvgIncome, Income_pct_AMI,  TotValue_1314, TotValue_1819, pct_change, newpropval)

data_clean <- data %>%
  filter(pct_change != "NA") %>%
  filter(pct_change != "NaN") %>%
  filter(pct_change != "Inf") %>%
  filter(ZIPCode != "NA") %>%
  arrange(desc(ZIPCode))
```

## Produce Table

```{r}
kable(head(data_clean))
```

This dataset includes 61 observations. 

# Descriptive Statistics

## Continuous Variables

The continuous variables we want to use are income, income percent of AMI, and percent change of property value.

### Sample Mean and 95-percent Confidence Interval
```{r}
income_t_test <- t.test(data_clean$AvgIncome)
incAMI_t_test <- t.test(data_clean$Income_pct_AMI)
prop_t_test <- t.test(data_clean$pct_change)
```

### Median and Interquartile Range
```{r}
income_quartiles <- quantile(data_clean$AvgIncome, na.rm = TRUE)
incAMI_quartiles <- quantile(data_clean$Income_pct_AMI, na.rm = TRUE)
prop_quartiles <- quantile(data_clean$pct_change, na.rm = TRUE)
```

### Standard Deviation
```{r}
income_st_dev <- sd(data_clean$AvgIncome, na.rm = TRUE)
incAMI_st_dev <- sd(data_clean$Income_pct_AMI, na.rm = TRUE)
prop_st_dev <- sd(data_clean$pct_change, na.rm = TRUE)
```

### Histograms
```{r}
income_hist <- ggplot(data_clean) +
  geom_histogram(aes(x = AvgIncome),
                 bins = 30)

incAMI_hist <- ggplot(data_clean) +
  geom_histogram(aes(x = Income_pct_AMI),
                 bins = 30)
  #scale_x_continuous(trans = "log")

prop_hist <- ggplot(data_clean) +
  geom_histogram(aes(x = pct_change),
                 bins = 100) + 
  scale_x_continuous(trans = "log")
```

## Displaying Results

### Formatted Table

```{r}
cont_summary <- tibble(
  Variable = c("Average Income", 
               "Income as percent of AMI", 
               "Change of property value (%)"),
  `Sample mean` = c(income_t_test$estimate,
                    incAMI_t_test$estimate,
                    prop_t_test$estimate),
  `Population mean (95% confidence) - low` = 
    c(income_t_test$conf.int[1],
      incAMI_t_test$conf.int[1],
      prop_t_test$conf.int[1]),
  `Population mean (95% confidence) - high` =
    c(income_t_test$conf.int[2],
      incAMI_t_test$conf.int[2],
      prop_t_test$conf.int[2]),
  Median = c(income_quartiles[3],
             incAMI_quartiles[3],
             prop_quartiles[3]),
  `Interquartile range` = c(income_quartiles[4] - income_quartiles[2],
                            incAMI_quartiles[4] - incAMI_quartiles[2],
                            prop_quartiles[4] - prop_quartiles[2]),
  `Standard deviation` = c(income_st_dev,
                          incAMI_st_dev,
                          prop_st_dev))

kable(cont_summary, digits = 0)
```
### Row of Histograms

```{r}
pretty_income_hist <- income_hist +
  theme_bw() +
  scale_x_continuous(name = "Median income") +
  scale_y_continuous(name = "Number of households living in affordable housing") +
  theme(axis.text.x = element_text(angle = 90))

pretty_incAMI_hist <- incAMI_hist +
  theme_bw() + 
  scale_x_continuous(name = "Income as percent of AMI",) +
  scale_y_continuous(name = "Number of households living in affordable housing") +
  theme(axis.text.x = element_text(angle = 90))

pretty_prop_hist = prop_hist +
  theme_bw() +
  scale_x_continuous(name = "Change in property value, as percent",
                     trans = "log",
                     breaks = c(0.1, 1, 10, 100, 1000, 10000),
                     labels = c("0.1", "1", "10", "100", "1000", "10,000")) +
  scale_y_continuous(name = "Number of households living in affordable housing") +
  theme(axis.text.x = element_text(angle = 90)) 

grid.arrange(pretty_income_hist, pretty_incAMI_hist, pretty_prop_hist,
             ncol = 3)
```

## Categorical Variables

We have two categorical variables:

* Whether the property value increased or decreased over the five-year period
* ZIP Code

```{r}
pct_inc <- t.test(data_clean$newpropval == "Increase")
pct_dec <-  t.test(data_clean$newpropval == "Decrease")

pct_94102 <- t.test(data_clean$ZIPCode == "94102")
pct_94103 <- t.test(data_clean$ZIPCode == "94103")
pct_94107 <- t.test(data_clean$ZIPCode == "94107")
pct_94108 <- t.test(data_clean$ZIPCode == "94108")
pct_94109 <- t.test(data_clean$ZIPCode == "94109")
pct_94110 <- t.test(data_clean$ZIPCode == "94110")
pct_94112 <- t.test(data_clean$ZIPCode == "94112")
pct_94114 <- t.test(data_clean$ZIPCode == "94114")
pct_94115 <- t.test(data_clean$ZIPCode == "94115")
pct_94117 <- t.test(data_clean$ZIPCode == "94117")
pct_94118 <- t.test(data_clean$ZIPCode == "94118")
pct_94123 <- t.test(data_clean$ZIPCode == "94123")
pct_94124 <- t.test(data_clean$ZIPCode == "94124")
pct_94125 <- t.test(data_clean$ZIPCode == "94125")
pct_94131 <- t.test(data_clean$ZIPCode == "94131")
pct_94133 <- t.test(data_clean$ZIPCode == "94133")
pct_94134 <- t.test(data_clean$ZIPCode == "94134")
```

```{r}
cat_summary_propval <- tibble(`Property value change` = 
                              c("Increase",
                                "Decrease"),
                            `Sample proportion` = 
                              c(pct_inc$estimate * 100,
                                pct_dec$estimate *100),
                            `95-percent confidence - low` = 
                              c(pct_inc$conf.int[1] * 100,
                                pct_dec$conf.int[1] * 100),
                            `95-percent confidence - high` = 
                              c(pct_inc$conf.int[2] * 100,
                                pct_dec$conf.int[2] * 100))

kable(cat_summary_propval, digits = 0)
```
### Displaying Results
```{r}
ggplot(cat_summary_propval) +
  geom_bar(aes(x = `Property value change`, 
               y = `Sample proportion`),
           stat = "identity") +
  geom_errorbar(aes(x = `Property value change`, 
               y = `Sample proportion`,
               ymin = `95-percent confidence - low`,
               ymax = `95-percent confidence - high`),
           stat = "identity") +
  scale_y_continuous(name = "% of households living in affordable housing",
                     breaks = c(0, 20, 40, 60, 80),
                     labels = c("0", "20%", "40%", "60%", "80%")) +
  theme_bw()
```

```{r}
cat_summary_ZIP <- tibble(`ZIP Codes` = 
                              c("94102",
                                "94103",
                                "94107",
                                "94108",
                                "94109",
                                "94110", 
                                "94112", 
                                "94114", 
                                "94115", 
                                "94117",
                                "94118",
                                "94123",
                                "94124",
                                "94125",
                                "94131",
                                "94133",
                                "94134"),
                            `Sample proportion` = 
                              c(pct_94102$estimate * 100,
                                pct_94103$estimate * 100,
                                pct_94107$estimate * 100,
                                pct_94108$estimate * 100,
                                pct_94109$estimate * 100, 
                                pct_94110$estimate * 100, 
                                pct_94112$estimate * 100, 
                                pct_94114$estimate * 100, 
                                pct_94115$estimate * 100, 
                                pct_94117$estimate * 100, 
                                pct_94118$estimate * 100,
                                pct_94123$estimate * 100, 
                                pct_94124$estimate * 100, 
                                pct_94125$estimate * 100, 
                                pct_94131$estimate * 100, 
                                pct_94133$estimate * 100, 
                                pct_94134$estimate * 100),
                            `95-percent confidence - low` = 
                              c(pct_94102$conf.int[1] * 100,
                                pct_94103$conf.int[1] * 100,
                                pct_94107$conf.int[1] * 100,
                                pct_94108$conf.int[1] * 100,
                                pct_94109$conf.int[1] * 100, 
                                pct_94110$conf.int[1] * 100, 
                                pct_94112$conf.int[1] * 100, 
                                pct_94114$conf.int[1] * 100, 
                                pct_94115$conf.int[1] * 100, 
                                pct_94117$conf.int[1] * 100, 
                                pct_94118$conf.int[1] * 100,
                                pct_94123$conf.int[1] * 100, 
                                pct_94124$conf.int[1] * 100, 
                                pct_94125$conf.int[1] * 100, 
                                pct_94131$conf.int[1] * 100, 
                                pct_94133$conf.int[1] * 100, 
                                pct_94134$conf.int[1] * 100),
                            `95-percent confidence - high` = 
                              c(pct_94102$conf.int[2] * 100,
                                pct_94103$conf.int[2] * 100,
                                pct_94107$conf.int[2] * 100,
                                pct_94108$conf.int[2] * 100,
                                pct_94109$conf.int[2] * 100, 
                                pct_94110$conf.int[2] * 100, 
                                pct_94112$conf.int[2] * 100, 
                                pct_94114$conf.int[2] * 100, 
                                pct_94115$conf.int[2] * 100, 
                                pct_94117$conf.int[2] * 100, 
                                pct_94118$conf.int[2] * 100,
                                pct_94123$conf.int[2] * 100, 
                                pct_94124$conf.int[2] * 100, 
                                pct_94125$conf.int[2] * 100, 
                                pct_94131$conf.int[2] * 100, 
                                pct_94133$conf.int[2] * 100, 
                                pct_94134$conf.int[2] * 100))

kable(cat_summary_ZIP, digits = 0)
```
```{r}
ggplot(cat_summary_ZIP) +
  geom_bar(aes(x = `ZIP Codes`, 
               y = `Sample proportion`),
           stat = "identity") +
  geom_errorbar(aes(x = `ZIP Codes`, 
               y = `Sample proportion`,
               ymin = `95-percent confidence - low`,
               ymax = `95-percent confidence - high`),
           stat = "identity") +
  scale_y_continuous(name = "% of households living in affordable housing",
                     breaks = c(0, 20, 40, 60, 80),
                     labels = c("0", "20%", "40%", "60%", "80%")) +
  theme_bw()
```


### References

* Center for Housing Policy. n.d. "'Don't Put it Here!' Does Affordable Housing Cause Nearby Property Values to Decline?" Insights from Housing Policy Research. https://furmancenter.org/files/media/Dont_Put_It_Here.pdf.

* Nguyen, Mai Thi. 2005. "Does Affordable Housing Detrimentally Affect Property Values? A Review of the Literature." Journal of Planning Literature 20, no. 1: 15-26. https://10.1177/0885412205277069

* PolicyMap. 2019. Location of publicly funded housing within San Francisco [Map].  https://harvard-policymap-com.ezp-prod1.hul.harvard.edu/maps?p=100385&o=101450,101500,96400,162200,150100,165200,306250&cx=-122.38390910801226&cy=37.80951024596563&cz=8

* San Francisco City & County, Office of Assessor-Recorder. n.d. Secured Property Tax Data. http://sfassessor.org/news-information/property-data-0

